---
title: "football insights"
author: "Victor T. Ngatunga"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
library(dplyr)
library(ggplot2)
library(caret)
library(readr)
library(randomForest)
```


```{r importing dataset}
 E0 <- read_csv("~/Downloads/E0.csv")
ECHO=FALSE

```
 
 Dataset- CSV english Premier League
```{r Creating match result variable}
E0$Result <- ifelse(E0$FTHG > E0$FTAG, 1,
             ifelse(E0$FTHG < E0$FTAG, -1, 0))
```
meaning 1:home win, 0:draw, -1: away win

```{r selecting features}
data <- E0 %>%
  select(HS, AS, HST, AST, Result) %>%
  na.omit()
ECHO=FALSE
```
HomeTeam, AwayTeam

FTHG – Full time home goals
FTAG – Full time away goals
HS, AS – total shots home, away
HST, AST – shots on target home, away

```{r training/ test split}
set.seed(123)

train_index <- createDataPartition(data$Result, p = 0.8, list = FALSE)

train <- data[train_index, ]
test  <- data[-train_index, ]
```

```{r training model}
model <- randomForest(as.factor(Result) ~ HS + AS + HST + AST,
                      data = train,
                      ntree = 500)

model
```

```{r Prediction basing on the model}
pred <- predict(model, test)
```

```{r Model evaluation}
confusionMatrix(pred, as.factor(test$Result))

```

```{r Feature importance}
varImpPlot(model)
```
shows what influence results the most.

```{r quick visualization}
ggplot(E0, aes(x = HS, y = FTHG)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "Home Shots vs Home Goals")
```

```{r }
ggplot(E0, aes(x=HST, y=FTHG, color=Result)) +
  geom_jitter(width=0.2, height=0.2, alpha=0.6) +
  geom_smooth(method="lm", se=FALSE, linetype="dotted") +
  labs(title="Home Shots on Target vs Home Goals by Match Outcome",
       x="Home Shots on Target", y="Home Goals") +
  theme_minimal()
```

Key Observations:

Precision beats volume: The scatter plot shows that matches with more shots on target (HST) tend to result in more goals for the home team, regardless of total shots.

Outcome clustering: Wins (colored points) are concentrated at higher shots-on-target values, while draws and losses cluster at lower values. This visually confirms the strong relationship between shot efficiency and match success.

Strategic insight: Focusing on converting shots into targets is a stronger driver for scoring than merely increasing shot attempts, highlighting a key performance metric for teams.

Recruiter-relevant takeaway: This visualization demonstrates the ability to turn raw match statistics into actionable insights, showing analytical thinking, attention to detail, and understanding of football data, exactly the skills required for a Sports Data Collector role.